Combinatorial Proofs and Decomposition Theorems for First-order Logic\vspace{-3ex}

### 2021-01-25 main.tex
tweak

A graph homomorphism $\phi\colon \tuple{\vG, \eG} \rightarrow
\tuple{\vH, \eH}$ is a \bfit{fibration} if for all $v \in \vG$ and
$w\phi(v) \in \eH$, there exists a unique $\tilde{w} \in \vG$ such
that $\tilde{w}v \in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-left),
and is a
\bfit{skew fibration} if for all $v \in \vG$ and $w\phi(v) \in \eH$
there exists $\tilde{w} \in \vG$ such that $\tilde{w}v \in \eG$ and
$\phi(\tilde{w})w \notin \eH$ (indicated below-centre).
A directed graph homomorphism is a
\bfit{fibration} if for all $v \in \vG$ and $\pair{w,\phi(v)} \in
\eH$, there exists a unique $\tilde{w} \in \vG$ such that $\pair{\tilde{w},v}
\in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-right).
%
\liftingdiagrams
%
A \bfit{fograph homomorphism} $\phi=\tuple{\phi,\rsubstof\phi}$ is a
pair where $\phi\colon\gG\to\gH$ is a graph homomorphism between the
underlying graphs, and $\rsubstof\phi$, also called the
\bfit{substitution induced by $\phi$} is a variable renaming 
such that
for all $v\in\vG$ we have
$\labelof{\phi(v)}=\rsubstof\phi(\labelof v)$,
and $\rsubstof\phi$ is the identity on variables not in $\gG$.
%
Note that $\phi$ necessarily maps binders to binders
and literals to literals.
%
Since $\rsubstof\phi$ is fully determined by $\phi$ alone, we often
leave $\rsubstof\phi$ implicit.
A fograph homomorphism $\phi\colon\gG \rightarrow \gH$
%%\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
% is \bfit{existential-preserving} 
\bfit{preserves existentials}
if for all existential binders $b$ in
$\gG$, the binder $\phi(b)$ is a existential in $\gH$.\dominic{tweaked this paragraph}
%
\begin{definition}
  Let $\gG$ and $\gH$ be fographs. A \bfit{skew bifibration}
  $\phi\colon\gG\to\gH$ is an existential-preserving fograph
  homomorphism that is a skew fibration on $\tuple{\vG, \eG} \to
  \tuple{\vH, \eH}$ and a fibration on the binding graphs $\bG\to\bH$.
  %etween fographs is a label-preserving and
%existential-preserving fograph homomorphism whose restriction to the
%binding graphs is a fibration.
\end{definition}
%
\begin{example}
  Below-left is a skew bifibration, whose binding fibration
  is below-centre. When the labels on the source fograph can be inferred
  (modulo renaming),
  we often omit the labeling in the upper graph, as below-right, which refer to as 
  \bfit{condensed form}.\dominic{I said ``can be inferred'' to account for 
  e.g.\
  $\graphof{\forall x_1\forall x_2 (px_i\vee px_j\vee px_k\vee px_l)} \to \forall x px$ whose
  unlabelled skeleton is the skeleton of many distinct formula variants (``who binds who?'').}
  \begin{center}\begin{math}
  \scalebox{.9}{
  \begin{tikzpicture}[graph]\drinkerfiblabelledpair{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerbindingfiblabelled{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerfib
  \end{tikzpicture}}
\end{math}\end{center}\end{example}
%
%%\subsection{First-order combinatorial proofs}
%
\begin{definition}
A \bfit{first-order combinatorial proof} (\bfit{FOCP}) of a fograph $\gG$ is a skew
bifibration $\phi\colon \gC \rightarrow \gG$ where $\gC$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof{A}$.
\end{definition}
%
Figure~\ref{fig:cps} shows examples of
FOCPs (taken from~\cite{hughes:fopws}), each above the formula it proves.
The same FOCPs are shown in
Figure~\ref{fig:cps-condensed} in condensed form.
%
\begin{thm}[\cite{hughes:fopws}]
  \label{thm:FOCP}
  FOCPs are sound and complete for first-order logic.
\end{thm}
%
\begin{remark}
  Our definition of FOCP is slightly more lax than the original
  definition of~\cite{hughes:fopws}, as we allow for a variable
  renaming $\substof\phi$ which was forced to be the identity
  in~\cite{hughes:fopws}.
  %The reason for this relaxation in our case is that we want to cover a richer class of proofs, also allowing 
\end{remark}

>>>>

A graph homomorphism $\phi\colon \tuple{\vG, \eG} \rightarrow
\tuple{\vH, \eH}$ is a \bfit{fibration} if for all $v \in \vG$ and
$w\phi(v) \in \eH$, there exists a unique $\tilde{w} \in \vG$ such
that $\tilde{w}v \in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-left),
and is a
\bfit{skew fibration} if for all $v \in \vG$ and $w\phi(v) \in \eH$
there exists $\tilde{w} \in \vG$ such that $\tilde{w}v \in \eG$ and
$\phi(\tilde{w})w \notin \eH$ (indicated below-centre).
A directed graph homomorphism is a
\bfit{fibration} if for all $v \in \vG$ and $\pair{w,\phi(v)} \in
\eH$, there exists a unique $\tilde{w} \in \vG$ such that $\pair{\tilde{w},v}
\in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-right).
%
\liftingdiagrams
%
A \bfit{fograph homomorphism} $\phi=\tuple{\phi,\rsubstof\phi}$ is a
pair where $\phi\colon\gG\to\gH$ is a graph homomorphism between the
underlying graphs, and $\rsubstof\phi$, also called the
\bfit{substitution induced by $\phi$} is a variable renaming 
such that
for all $v\in\vG$ we have
$\labelof{\phi(v)}=\rsubstof\phi(\labelof v)$,
and $\rsubstof\phi$ is the identity on variables not in $\gG$.
%
Note that $\phi$ necessarily maps binders to binders
and literals to literals.
%
Since $\rsubstof\phi$ is fully determined by $\phi$ alone, we often
leave $\rsubstof\phi$ implicit.
A fograph homomorphism $\phi\colon\gG \rightarrow \gH$
%%\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
% is \bfit{existential-preserving} 
\bfit{preserves existentials}
if for all existential binders $b$ in
$\gG$, the binder $\phi(b)$ is a existential in $\gH$.\dominic{tweaked this paragraph}
%
\begin{definition}
  Let $\gG$ and $\gH$ be fographs. A \bfit{skew bifibration}
  $\phi\colon\gG\to\gH$ is an existential-preserving fograph
  homomorphism that is a skew fibration on $\tuple{\vG, \eG} \to
  \tuple{\vH, \eH}$ and a fibration on the binding graphs $\bG\to\bH$.
  %etween fographs is a label-preserving and
%existential-preserving fograph homomorphism whose restriction to the
%binding graphs is a fibration.
\end{definition}
%
\begin{example}
  Below-left is a skew bifibration, whose binding fibration
  is below-centre. When the labels on the source fograph can be inferred
  (modulo renaming),
  we often omit the labeling in the upper graph, as below-right, which refer to as 
  \bfit{condensed form}.\dominic{I said ``can be inferred'' to account for 
  e.g.\
  $\graphof{\forall x_1\forall x_2 (px_i\vee px_j\vee px_k\vee px_l)}
  \to \graphof{\forall x\mkern2mu px}$ whose
  unlabelled skeleton is the skeleton of many distinct formula variants (``who binds who?'').}
  \begin{center}\begin{math}
  \scalebox{.9}{
  \begin{tikzpicture}[graph]\drinkerfiblabelledpair{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerbindingfiblabelled{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerfib
  \end{tikzpicture}}
\end{math}\end{center}\end{example}
%
%%\subsection{First-order combinatorial proofs}
%
\begin{definition}
A \bfit{first-order combinatorial proof} (\bfit{FOCP}) of a fograph $\gG$ is a skew
bifibration $\phi\colon \gC \rightarrow \gG$ where $\gC$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof{A}$.
\end{definition}
%
Figure~\ref{fig:cps} shows examples of
FOCPs (taken from~\cite{hughes:fopws}), each above the formula it proves.
The same FOCPs are shown in
Figure~\ref{fig:cps-condensed} in condensed form.
%
\begin{thm}[\cite{hughes:fopws}]
  \label{thm:FOCP}
  FOCPs are sound and complete for first-order logic.
\end{thm}
%
\begin{remark}
  Our definition of FOCP is slightly more lax than the original
  definition of~\cite{hughes:fopws}, as we allow for a variable
  renaming $\substof\phi$ which was forced to be the identity
  in~\cite{hughes:fopws}.
  %The reason for this relaxation in our case is that we want to cover a richer class of proofs, also allowing 
\end{remark}
